---
output: github_document
---

## Introduction 

The goal of this project is to explore a dataset of tech industry job listings to better understand what drives salary differences across roles, locations, and qualifications. The technology sector is one of the fastest-growing industries worldwide, and compensation varies enormously depending on a wide range of factors. Analysis on this topic can help job seekers benchmark their earnings, guide career decisions, and reveal which qualifications translate most directly into higher pay. Ultimately, our goal is to understand the circumstances under which tech professionals earn more — and to build a predictive model that can estimate salary given a set of job attributes.

In pursuit of the stated goal, we will explore the following questions:

1. Which job titles command the highest salaries? Are AI and machine learning roles outpacing traditional software engineering positions?

2. How does experience level shape compensation? Is the relationship linear, or are there inflection points where salary growth accelerates?

3. Does education level significantly impact salary? What is the real-world premium for a PhD versus a Bachelor's degree?

4. How does geography affect pay? Which locations offer the highest compensation, and does fully remote work close the gap?

5. Does company size determine salary outcomes? Are large enterprises or startups more generous to tech workers?

6. Do the number of skills and certifications a worker holds translate into measurably higher pay?
    
7. Does the salary premium of having a higher education degree depend on how much experience you have? Reorder the levels of education_level from least to most educated and exclude all High School workers from the rest of this question. How many records remain? Create a new variable that groups workers into "Junior" (0–7 years) and "Senior" (8–20 years). Find a plot that shows how the relationship between education level and salary differs between junior and senior workers, and comment on what you find.


8. Are AI and Machine Learning roles actually earning more than traditional software roles, and does that gap hold across all locations?

These are the main questions we are looking to answer through the completion of this project. With the findings we will be able to draw meaningful conclusions on tech compensation, and ultimately build a machine learning model capable of predicting salary from a worker's profile.


## Data

## Structure

The dataset used in this project is the Job Salary Prediction Dataset, sourced from Kaggle. It contains 250,000 rows, each representing a unique job record, and 10 variables covering the worker's role, qualifications, employer characteristics, and annual salary in USD. The dataset spans 12 distinct job titles across 10 industries and 10 geographic locations, making it a broad cross-section of the global tech labor market.

The variables in the dataset are as follows. `job_title` records the employee's role such as AI Engineer or Data Analyst. `experience_years` is a numeric variable ranging from 0 to 20 years. `education_level` is an ordinal variable with five levels: High School, Diploma, Bachelor, Master, and PhD. `skills_count` records the number of technical skills the employee lists, ranging from 1 to 19. `industry` captures the sector the employer operates in, such as Finance, Healthcare, or Technology. `company_size` is an ordinal variable with five levels: Startup, Small, Medium, Large, and Enterprise. `location` records the country where the role is based, or "Remote" if the worker has no fixed location. `remote_work` specifies whether the arrangement is fully remote, hybrid, or on-site. `certifications` counts the number of professional certifications the worker holds, from 0 to 5. Finally, `salary` is the target variable — the annual salary in US dollars.

### Cleaning

First, we read in the dataset and inspected its structure.

```{r}
library(tidyverse)
library(caret)
library(randomForest)
library(xgboost)
library(Metrics)

df <- read_csv("job_salary_prediction_dataset.csv")
glimpse(df)
```

We then checked for missing values across all columns.

``` {r}
# check missing values 

colSums(is.na(df))
```

No missing values were found in any column. The dataset was entirely complete, requiring no imputation or row removal. We next reviewed the distribution of the target variable, `salary`.

``` {r}
# Get summary statistics for the target variable (salary)

summary(df$salary)
```


The salary ranges from a minimum of $31,867 to a maximum of $333,046, with a mean of $145,718 and a median of $143,453. The proximity of the mean and median suggests that the distribution is relatively symmetric, with no severe skew caused by extreme outliers. 

We then converted the categorical string columns into factors, applying an ordered factor structure to `education_level` and `company_size` to preserve their natural ranking.

``` {r}
# Convert categorical columns to factors, apply ordering where categories have a natural rank

df <- df %>%
  mutate(
    job_title       = as.factor(job_title),
    education_level = factor(education_level,
                             levels = c("High School", "Diploma", "Bachelor",
                                        "Master", "PhD"),
                             ordered = TRUE),
    industry        = as.factor(industry),
    company_size    = factor(company_size,
                             levels = c("Startup", "Small", "Medium",
                                        "Large", "Enterprise"),
                             ordered = TRUE),
    location        = as.factor(location),
    remote_work     = as.factor(remote_work)
  )
```

With the data cleaned and typed correctly, we proceeded to exploratory analysis and model building.

## Analysis

### Which job titles command the highest salaries?

#### Salary by Job Title

``` {r}
# Calculate mean salary per job title and plot as a horizontal bar chart, sorted highest to lowest
df %>%
  group_by(job_title) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(desc(mean_salary)) %>%
  ggplot(aes(x = reorder(job_title, mean_salary), y = mean_salary)) +
  geom_col(fill = "#2b6cb0") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Job Title") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Job Title")
```
```{r}
df %>%
  group_by(job_title) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(desc(mean_salary))
```

This bar chart shows the mean annual salary for each of the 12 job titles in the dataset, sorted from highest to lowest. AI Engineer is the top-earning role at $173,498 on average, followed by Machine Learning Engineer at $163,023 and Product Manager at $157,595. At the lower end, Business Analyst averages $122,551 and Data Analyst averages $119,892. The gap between the highest and lowest-paid roles is over $53,000, illustrating how dramatically job title alone can affect compensation within the same industry. The strong performance of AI and machine learning roles reflects the high demand for these specializations in the current market.

#### Salary Distribution by Job Title

``` {r}
# Plot the full salary distribution for each job title using boxplots, ordered by median salary

df %>%
  ggplot(aes(x = reorder(job_title, salary, FUN = median), y = salary)) +
  geom_boxplot(fill = "#e8f0fb", color = "#2b6cb0") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Job Title") + ylab("Salary (USD)") +
  ggtitle("Salary Distribution by Job Title")
```


The boxplot view reveals not just the averages but the spread within each role. All 12 job titles show a wide interquartile range, suggesting that salary is not solely determined by title. Even Data Analysts, the lowest-paid group on average, have individuals earning over $200,000. This indicates that other factors — experience, location, company size — contribute heavily to individual outcomes within each role.

### How does experience level shape compensation?

#### Experience and Salary

``` {r}
# Calculate mean salary for each experience year and plot as a line chart to show salary growth over time

df %>%
  group_by(experience_years) %>%
  summarize(mean_salary = mean(salary)) %>%
  ggplot(aes(x = experience_years, y = mean_salary)) +
  geom_line(color = "#c84b11", linewidth = 1.2) +
  geom_point(color = "#c84b11", size = 2.5) +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Years of Experience") + ylab("Mean Salary (USD)") +
  ggtitle("Mean Salary by Years of Experience")
```


The line chart shows a clear and consistent positive relationship between experience and salary. Workers with zero years of experience earn an average of $118,873, while those with 20 years earn $173,180, a difference of $54,307 over a full career. The growth is nearly linear, with no single dramatic jump, suggesting that salary increases accumulate steadily with each additional year rather than at specific career milestones. 

```{r}
# Calculate the correlation between experience years and salary
cor(df$experience_years, df$salary)
```

The correlation between `experience_years` and `salary` is 0.44, making it the strongest numeric predictor in the dataset.

#### Experience and Salary by Job Title

``` {r}
# Plot salary growth by experience for selected job titles using a line chart

df %>%
  filter(job_title %in% c("AI Engineer", "Data Analyst",
                           "Software Engineer", "Product Manager")) %>%
  group_by(job_title, experience_years) %>%
  summarize(mean_salary = mean(salary), .groups = "drop") %>%
  ggplot(aes(x = experience_years, y = mean_salary, color = job_title)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Years of Experience") + ylab("Mean Salary (USD)") +
  ggtitle("Salary Growth by Experience for Selected Roles") +
  labs(color = "Job Title")
```

The line chart breaks down salary growth by experience for four selected job titles. AI Engineer is the highest-paying role throughout, starting at approximately $150,000 with no experience and reaching close to $190,000 at 20 years. Product Manager follows closely behind, while Software Engineer sits in the middle range. Data Analyst is the lowest-paid role across all experience levels, starting around $90,000 and reaching approximately $145,000 at 20 years. Importantly, all four lines grow at a roughly similar rate — the gap between roles remains relatively constant rather than widening over time. This suggests that experience rewards all job titles fairly equally, and that the salary premium of choosing a higher-paying role like AI Engineer over Data Analyst is present from the very start of a career rather than accumulating over time.

```{r}
# Compare correlations of all numeric predictors with salary
cor(df[, c("experience_years", "skills_count", "certifications", "salary")])
```

Among all numeric predictors, experience_years has the strongest correlation with salary at 0.44, followed by skills_count at 0.13 and certifications at 0.07. This confirms that experience is the most important numeric driver of compensation, while skills and certifications have relatively little impact on their own.

### Does education level significantly impact salary?

#### Education Level and Salary

``` {r}
# Calculate mean salary per education level and plot as a bar chart

df %>%
  group_by(education_level) %>%
  summarize(mean_salary = mean(salary)) %>%
  ggplot(aes(x = education_level, y = mean_salary)) +
  geom_col(fill = "#276749") +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Education Level") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Education Level")
```

```{r}
df %>%
  group_by(education_level) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(education_level)
```

The bar chart shows a clear positive relationship between educational attainment and salary. High School graduates earn an average of $131,715, while PhD holders earn $163,976 — a premium of $32,261. The incremental gains between each level are: High School to Diploma $5,444; Diploma to Bachelor $5,252; Bachelor to Master $10,894; and Master to PhD $10,671. The larger jumps at the graduate level suggest advanced degrees carry a more meaningful salary premium. However, the full education premium of $32,261 is smaller than the $54,307 gained through 20 years of experience, suggesting experience is a stronger salary driver than education in this dataset.

``` {r}
# Plot the full salary distribution for each education level using boxplots

df %>%
  ggplot(aes(x = education_level, y = salary)) +
  geom_boxplot(fill = "#e8f5ee", color = "#276749") +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Education Level") + ylab("Salary (USD)") +
  ggtitle("Salary Distribution by Education Level")
```

The boxplot reveals substantial overlap across all education levels, confirming that education alone does not determine salary. While higher degrees shift the distribution upward, a Bachelor's holder can still out-earn a PhD depending on experience, role, and location.

### How does geography affect pay?

#### Salary by Location

``` {r}
# Calculate mean salary per location and plot as a horizontal bar chart, sorted highest to lowest

df %>%
  group_by(location) %>%
  summarize(mean_salary = mean(salary), n = n()) %>%
  arrange(desc(mean_salary)) %>%
  ggplot(aes(x = reorder(location, mean_salary), y = mean_salary)) +
  geom_col(fill = "#5a3e8a") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Location") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Location")
```

```{r}
df %>%
  group_by(location) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(desc(mean_salary))
```

Geography is one of the most powerful salary determinants in the dataset. Workers in the USA earn $181,716 on average — the highest of any location. Canada is second at $167,391, followed by the UK at $160,075 and Germany at $153,376. Remote, Sweden, Australia, Singapore, and Netherlands all cluster tightly between $139,294 and $139,443, suggesting these markets are similarly priced. India sits significantly below all other locations at $97,690, reflecting cost-of-living differences and local market structures. The gap between the USA and India is $84,026 — making location the single largest salary differentiator in the dataset.

#### Remote Work and Salary

``` {r}
# Calculate mean salary by remote work status and plot as a bar chart

df %>%
  group_by(remote_work) %>%
  summarize(mean_salary = mean(salary)) %>%
  ggplot(aes(x = remote_work, y = mean_salary)) +
  geom_col(fill = "#5a3e8a") +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Remote Work Status") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Remote Work Status")
```

```{r}
df %>%
  group_by(remote_work) %>%
  summarize(mean_salary = mean(salary))
```

Fully remote workers earn a mean salary of $149,280, which is $5,348 higher than hybrid workers ($143,970) and $5,348 higher than on-site workers ($143,932). The hybrid and on-site averages are nearly identical, suggesting that requiring physical presence — whether full or part-time — has no additional wage premium. The slight advantage for fully remote workers may reflect that remote roles attract applicants from higher-paying markets, or that companies offering full remote flexibility compensate with higher base salaries.

### Q5. Does company size determine salary outcomes?

#### Company Size and Salary

``` {r}
# Calculate mean salary per company size and plot as a bar chart
df %>%
  group_by(company_size) %>%
  summarize(mean_salary = mean(salary)) %>%
  ggplot(aes(x = company_size, y = mean_salary)) +
  geom_col(fill = "#b7791f") +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Company Size") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Company Size")
```
```{r}
df %>%
  group_by(company_size) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(company_size)
```

There is a strong and monotonic relationship between company size and salary. Startups pay an average of $127,289 — the lowest of all five categories. As company size grows, salary increases consistently: Small companies pay $134,357, Medium pay $141,538, Large pay $155,711, and Enterprise companies pay the most at $169,616. The gap between Startup and Enterprise is $42,327, suggesting that larger organizations have more financial resources and structured compensation bands to attract senior talent.

### Q6. Do skills and certifications translate into higher pay?


#### Skills Count and Salary

``` {r}
# Calculate mean salary per skills count and plot as a line chart

df %>%
  group_by(skills_count) %>%
  summarize(mean_salary = mean(salary)) %>%
  ggplot(aes(x = skills_count, y = mean_salary)) +
  geom_line(color = "#c84b11", linewidth = 1) +
  geom_point(color = "#c84b11", size = 2) +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Number of Skills") + ylab("Mean Salary (USD)") +
  ggtitle("Mean Salary by Skills Count")
```

```{r}
df %>%
  group_by(skills_count) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(skills_count)
```

The line chart shows a positive relationship between skills count and salary. Workers with 1 skill earn an average of $137,788, while those with 19 skills earn $153,156 — a difference of $15,368. The growth is steady but moderate, consistent with the weak correlation of 0.13 observed earlier.

#### Certifications and Salary

``` {r}
# Calculate mean salary per certification count and plot as a bar chart
df %>%
  group_by(certifications) %>%
  summarize(mean_salary = mean(salary)) %>%
  ggplot(aes(x = factor(certifications), y = mean_salary)) +
  geom_col(fill = "#2b6cb0") +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Number of Certifications") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Number of Certifications")
```


Certifications follow a consistent stair-step pattern. Workers with zero certifications earn $141,492 on average, while those with five certifications earn $149,607 — a gain of $8,115. Each additional certification adds between $1,200 and $2,000 to average salary. The relationship is smooth and monotonic with no reversals, suggesting that certifications are reliably associated with higher pay. The correlation between `certifications` and `salary` is 0.07, making it the weakest numeric predictor in the dataset, but the consistent direction of the effect is notable.

#### Industry and Salary

``` {r}
# Calculate mean salary per industry, zoom in using coord_cartesian to show differences clearly

df %>%
  group_by(industry) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(desc(mean_salary)) %>%
  ggplot(aes(x = reorder(industry, mean_salary), y = mean_salary)) +
  geom_col(fill = "#718096") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format()) +
  coord_cartesian(ylim = c(145000, 146500)) +
  xlab("Industry") + ylab("Mean Salary (USD)") +
  ggtitle("Average Salary by Industry")
```
```{r}
df %>%
  group_by(industry) %>%
  summarize(mean_salary = mean(salary)) %>%
  arrange(desc(mean_salary))
```

When zoomed in, the chart shows that all 10 industries fall within a very narrow range of just $594 — from Retail at $145,400 to Education at $145,994. While differences exist, they are negligibly small compared to other variables like location ($84,026 gap) or job title ($53,000 gap). Industry is effectively the weakest salary predictor in this dataset.

## Salary Prediction Model

### Train / Test Split

To evaluate our model on unseen data, we split the dataset into a
training set (80%) and a test set (20%).

``` {r}
# Split data into 80% training and 20% test set using createDataPartition
set.seed(42)
train_index <- createDataPartition(df$salary, p = 0.80, list = FALSE)
train_data  <- df[ train_index, ]
test_data   <- df[-train_index, ]

nrow(train_data)
```

``` {r}
# Confirm the number of rows in the test set
nrow(test_data)
```

The dataset was split into 80% training (200,001 rows) and 20% test (49,999 rows). The model trains on the training set and is evaluated on the unseen test set to measure real prediction accuracy.

### Model 1: Linear Regression

We began with a linear regression model as our baseline. Linear regression fits a straight-line relationship between each predictor and the target variable. Before fitting the model, we one-hot encoded the factor variables into binary columns using `dummyVars`.

``` {r}
# One-hot encode factor variables, fit linear regression model, 
# and evaluate using RMSE and MAE on test set

dummy_model   <- dummyVars(salary ~ ., data = train_data, fullRank = TRUE)
train_encoded <- predict(dummy_model, newdata = train_data) %>% as.data.frame()
test_encoded  <- predict(dummy_model, newdata = test_data)  %>% as.data.frame()

train_encoded$salary <- train_data$salary
test_encoded$salary  <- test_data$salary

lm_model <- lm(salary ~ ., data = train_encoded)
lm_preds <- predict(lm_model, newdata = test_encoded)

cat("RMSE:", round(rmse(test_data$salary, lm_preds), 0), "\n")
cat("MAE: ", round(mae(test_data$salary,  lm_preds), 0), "\n")
```

The linear regression model produces an RMSE of $7,149 and an MAE of $5,450. This means on average the model's salary predictions are off by roughly $5,450. While this establishes a usable baseline, linear regression assumes straight-line relationships between predictors and salary, ignoring interactions and non-linear effects

### Model 2: XGBoost

XGBoost (Extreme Gradient Boosting) builds trees sequentially, with
each new tree correcting the errors of the previous ones. It is often
the most accurate algorithm for tabular data. XGBoost requires all
inputs to be numeric, so we encoded factors as integers before fitting.
```{r}
library(tidyverse)
library(xgboost)
library(Metrics)
```

``` {r}
# Encode factors as integers, build XGBoost model, and evaluate on test set
encode_for_xgb <- function(data) {
  data %>%
    mutate(across(where(is.factor), as.integer)) %>%
    select(-salary) %>%
    as.matrix()
}

X_train <- encode_for_xgb(train_data)
X_test  <- encode_for_xgb(test_data)

dtrain <- xgb.DMatrix(data = X_train, label = train_data$salary)
dtest  <- xgb.DMatrix(data = X_test,  label = test_data$salary)

xgb_params <- list(
  objective        = "reg:squarederror",
  eta              = 0.1,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

set.seed(42)
xgb_model <- xgb.train(
  params        = xgb_params,
  data          = dtrain,
  nrounds       = 300,
  evals     = list(train = dtrain, test = dtest),
  verbose       = 1,
  print_every_n = 100
)
```

``` {r}
xgb_preds <- predict(xgb_model, dtest)

cat("RMSE:", round(rmse(test_data$salary, xgb_preds), 0), "\n")
cat("MAE: ", round(mae(test_data$salary,  xgb_preds), 0), "\n")
```

XGBoost achieves the best performance of the two models, with an RMSE of $5,201 and an MAE of $4,146. The training log shows RMSE dropping from $35,249 at round 1 to $5,061 at round 300, confirming the model improves steadily with each boosting round. We then inspected which features the model relied on most heavily.

``` {r}
# Plot top 10 most important features from XGBoost model
xgb_importance <- xgb.importance(model = xgb_model)
xgb.plot.importance(xgb_importance, top_n = 10,
                    main = "XGBoost: Feature Importance (Gain)")
```


Location is the most important predictor by a wide margin, followed by experience_years and company_size. Industry and remote_work contribute the least, consistent with our earlier findings.

### Model Comparison

``` {r}
# Compare RMSE and MAE across both models
results <- data.frame(
  Model = c("Linear Regression", "XGBoost"),
  RMSE  = c(7149, 5201),
  MAE   = c(5450, 4146)
)
print(results)
```

XGBoost outperforms Linear Regression on both metrics. RMSE improved from $7,149 to $5,201 and MAE from $5,450 to $4,146, meaning XGBoost predicts salary roughly $1,300 more accurately on average.

``` {r}
# Visualize RMSE and MAE for both models as a side-by-side bar chart

results %>%
  pivot_longer(cols = c(RMSE, MAE), names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = reorder(Model, Value), y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab(NULL) + ylab("Error (USD)") +
  ggtitle("Model Comparison: RMSE and MAE (lower is better)")
```

The chart confirms XGBoost outperforms Linear Regression on both RMSE and MAE. Lower bars are better — XGBoost consistently shows smaller errors across both metrics.

### Residual Analysis

To evaluate whether the best model makes systematic errors, we examined the distribution of residuals — the differences between actual and predicted salaries.

``` {r}
# The distribution of residuals — the differences between actual and predicted salaries...

residuals_df <- data.frame(
  actual    = test_data$salary,
  predicted = xgb_preds,
  residual  = test_data$salary - xgb_preds
)

ggplot(residuals_df, aes(x = predicted, y = actual)) +
  geom_point(alpha = 0.05, color = "#2b6cb0", size = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "#c84b11", linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Predicted Salary") + ylab("Actual Salary") +
  ggtitle("Predicted vs. Actual Salary (XGBoost)")
```


The predicted vs. actual plot shows predictions clustering tightly and consistently along the diagonal across the full salary range. The error remains stable from low to high earners with no widening at the extremes, confirming that XGBoost performs equally well across all salary levels.

``` {r}
# The residuals form a tight, symmetric bell centered at zero.

ggplot(residuals_df, aes(x = residual)) +
  geom_histogram(bins = 80, fill = "#276749", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "#c84b11", linewidth = 1, linetype = "dashed") +
  scale_x_continuous(labels = scales::dollar_format()) +
  xlab("Residual (Actual - Predicted)") + ylab("Count") +
  ggtitle("Distribution of Residuals (XGBoost)")
```

The residual distribution is approximately bell-shaped and centered near zero, which is a desirable property. There are no dramatic long tails or asymmetry, indicating that the model does not systematically over- or under-predict salary for any particular segment of the data. The spread of roughly ±$20,000 is consistent with the reported RMSE.

### Predicting a New Salary

Once trained, the XGBoost model can be used to estimate the salary for any new combination of job attributes. Below is an example prediction for a hypothetical AI Engineer with 8 years of experience, a Master's degree, 12 skills, and 3 certifications, working at a Large company in the USA in a Hybrid arrangement.

``` {r}
new_employee <- data.frame(
  job_title        = factor("AI Engineer",    levels = levels(df$job_title)),
  experience_years = 8,
  education_level  = factor("Master",
                             levels = levels(df$education_level),
                             ordered = TRUE),
  skills_count     = 12,
  industry         = factor("Technology",     levels = levels(df$industry)),
  company_size     = factor("Large",
                             levels = levels(df$company_size),
                             ordered = TRUE),
  location         = factor("USA",            levels = levels(df$location)),
  remote_work      = factor("Hybrid",         levels = levels(df$remote_work)),
  certifications   = 3
)

new_matrix <- encode_for_xgb(cbind(new_employee, salary = 0))
new_pred   <- predict(xgb_model, xgb.DMatrix(new_matrix))

cat("Predicted Salary: $", round(new_pred, 0), "\n")
```


### Question 7
```{r}
# Reorder education_level from least to most educated
df$education_level <- factor(df$education_level,
                             levels = c("High School", "Diploma", "Bachelor",
                                        "Master", "PhD"),
                             ordered = TRUE)

# Exclude High School workers
df2 <- df %>% filter(education_level != "High School")
nrow(df2)

# Create Junior / Senior variable
df2 <- df2 %>%
  mutate(exp_group = ifelse(experience_years <= 7, "Junior", "Senior"),
         exp_group = factor(exp_group, levels = c("Junior", "Senior")))

# Does education pay off more when you're senior vs junior?
ggplot(df2, aes(x = education_level, y = salary, fill = education_level)) +
  geom_boxplot(outlier.alpha = 0.2) +
  facet_wrap(~ exp_group) +
  scale_y_continuous(labels = scales::dollar_format()) +
  xlab("Education Level") +
  ylab("Salary (USD)") +
  labs(fill = "Education Level") +
  ggtitle("Does Education Pay Off More for Senior Workers?") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

After excluding High School workers, the faceted boxplot shows that education pays off more for Senior workers than Junior ones. For Junior workers (0–7 years), salary differences across education levels are relatively small. For Senior workers (8–20 years), the gap between Diploma and PhD is noticeably wider, with PhD holders earning significantly more. This suggests that advanced degrees become more valuable as workers gain experience — the combination of high education and seniority yields the highest salaries.

### QUESTION 8 
```{r}

df %>%
  group_by(location, job_title) %>%
  summarise(mean_salary = mean(salary), .groups = "drop") %>%
  mutate(job_title = fct_reorder(job_title, mean_salary)) %>%
  ggplot(aes(x = mean_salary, y = job_title, color = location)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Mean Salary by Job Title and Location",
       x = "Mean Salary (USD)", y = NULL, color = "Location") +
  theme_minimal()
```

The dot plot confirms that AI Engineer and Machine Learning Engineer consistently earn the most across almost all locations, while Data Analyst and Business Analyst remain at the bottom. The USA consistently has the highest dots for every job title, while India has the lowest. Notably, the gap between locations is visible across all roles meaning the location premium holds regardless of job title. This confirms that both job title and location independently influence salary, and their effects stack.

## Conclusion.

In conclusion, our analysis of 250,000 global tech job records and subsequent machine learning models have shed light on several important drivers of salary in the technology industry. The most consistent finding throughout our analysis was that salary is heavily influenced by a small number of factors, while others — most notably industry — have almost no impact at all.

Job title is the most intuitive salary driver, with AI and machine learning roles commanding a premium of $50,000 or more over traditional analyst roles. Experience accumulates steadily and meaningfully, contributing roughly $54,000 across a 20-year career. Geography introduces the starkest divide of all, with US-based workers earning nearly double their Indian counterparts in the same roles. Company size matters significantly, with Enterprise employers paying $42,000 more than Startups on average. Education contributes a more modest premium, with PhD holders earning around $32,000 more than High School graduates. Skills and certifications are positively correlated with pay but explain relatively little of the overall variance. And industry, despite being a variable many job seekers consider carefully, turns out to be essentially irrelevant within the tech sector.

In terms of prediction, XGBoost outperformed linear regression, achieving an MAE of $4,146 on the held-out test set. This means that for most workers, the model can estimate annual salary within roughly $19,000, a useful tool for benchmarking, but still subject to the noise introduced by factors not captured in the dataset, such as individual negotiation, equity compensation, and company-specific pay bands.

Further research could explore interaction effects — for example, whether the experience premium is larger for AI Engineers than for Data Analysts — and whether the location effect is stable across all job titles or concentrated in certain roles. Incorporating additional variables such as programming language, years at current company, and total compensation (including equity) would also improve the model's practical utility.
